---
title: "Dyngen data generation pipeline"
output: html_notebook
---


```{r setup, include=FALSE}
# knitr::opts_chunk$set(fig.path = "man/figures/README_", warning=FALSE, message=FALSE, error=FALSE, echo = TRUE)
rm(list = ls())
gc()
```

```{r}
library(tidyverse)
library(rlang)
library(dyngen)
library(dyngen.manuscript)
```
A package for generating synthetic single-cell data starting from regulatory networks. The data is generated in several steps:


## Just two steps-simplified ver.

### Step 1: Initialize the model and set the parameters
**The most important step**

A dyngen simulation can be started by providing a backbone to the `initialise_model()` function.
The backbone of a `dyngen` model is what determines the overall dynamic process 
that a cell will undergo during a simulation. It consists of a set of gene modules, which regulate
eachother in such a way that expression of certain genes change over time in a specific manner.

```
model <- initialise_model(backbone, num_cells, num_tfs, num_targets, num_hks, distance_matric,tf_network_params, feature_network_params, kinetics_params, gold_standard_params, simulation_params, experiment_params, verbose, download_cache_dir, num_cores, id )
```

- `backbone` the backbone to choose, we can either do it by 

```{r}
backbones <- list_backbones()
backbones_name <- names(backbones)

backb <- backbones[[backbones_name[1]]]
# or 
# backb <- backbones$bifurcating()
```

or do it by, there are corresponding functions, returning a list with elements named as `module_info`, `module_network` and `expression_patterns` 

```{r}
backb <- backbone_bifurcating()
attributes(backb)

```
```{r}
?initialise_model()
```
- `num_cells`, default `1000`, number of cells or samples
- `num_tfs`, the number of transcription factors (TFs) to generate. TFs can only be controled by other TFs
- `num_targets`, the number of target genes to generate, target genes can only be regulated by other target genes or TFs
- `num_hks`, the number of housekeeping genes to generate, HKs are highly expressed but not regulated by TFs and targets
- `distance_metric`, the distance matric to be used to calculate the distance between cells, note that aside from `euclidean`, we have `pearson`, `spearman` 
```{r}
?tf_network_default()
```
- 
```
tf_network_params = tf_network_default(min_tfs_per_modules = 1L, sample_num_regulators = function() 2, weighted_sampling = FALSE)
```
generate a network for tfs, basically determines the number of tfs for each module, and the number of tfs a tfs can control (function),  
```{r}
?kinetics_default()
```
- 
```
feature_network_params = feature_network_default(realnet = NULL, damping = 0.01, target_resampling = Inf, max_in_degree = 5)
``` 
generate a network for target genes that are regulated by the tfs and housekeeping genes, `max_in_degree` is maximal indegree of each target/HKs. `realnet` the name of GRN in realnets, or sampling randomly from realnet, purpose is to generate the target/HKs. `damping` parameter for realnet sampling process. Something is not usually changed 

**Kinetic parameter for data generation**
- `kinetics_params = kinetics_default(sampler_tfs(), sampler_nontfs, sampler_interactions())` configure the params for `generate_kinetics()` function.

To write different kinetics settings, you need to rewrite three functions `sampler_tfs()`, `sampler_nontfs()`, `sampler_interations()`, all three functions should be wrote with the setting `function(feature_info, feature_network, cache_dir = NULL, verbose = FALSE)`

```{r}
# parameters sampled from distribution
sampler_tfs <- function(feature_info, feature_network, cache_dir = NULL, verbose = FALSE){
  feature_info %>% 
    mutate(transcription_rate = transcription_rate %|%
             runif(n(), 1, 2),
           translation_rate = translation_rate %|%
             runif(n(), 100, 150),
           mrna_halflife = mrna_halflife %|%
             runif(n(), 2.5, 5),
           protein_halflife=protein_halflife %|%
             runif(n(), 5, 10),
           independence = independence %|% 1,
           splicing_rate = splicing_rate %|% (log(2)/(10/60)))
}

# details pls check the function below
# format(kinetics_default())
```

**example in dyngen manuscript**
```{r}
kinetic_params <- kinetics_default()
kinetic_params$sampler_tfs <- function(...){
  x <- kinetics_default()$sampler_tfs(...)
  x$splicing_rate <- x$splicing_rate * spling_multiplier
  # more modifications...
  x
}
```

`sampler_nontfs()` samples the `transcription_rate`, `translation_rate`, `mrna_halflife` and `protein_halflife` from a supplementary file of Schwannhäusser et al., 2011, doi.org/10.1038/nature10098. **splicing_rate is by default the same as in sampler_tfs()**. independence is sampled from U(0, 1).

- `gold_standard_params = gold_standard_default(tau = 30/3600, census_interval = 10/60)` parameter setting for `generate_gold_standard()`

`tau = 30/3600` the time step of the ODE algorithm to generate the gold standard

`census_interval = 10/60` a granularity parameter of the gold standard time steps, larger or equal to `tau`
```{r}
?simulation_default()
```

**simulation parameters setting**
- `simulation_params = simulation_default()` 
```
simulation_default(
  # NULL then infered from backbone
  burn_time = NULL,
  
  # if NULL then infered from backbone
  total_time = NULL,
  
  # params for GillespieSSA2::ssa()
  ssa_algorithm = ssa_etl(tau = 30/3600),
  
  # granularity parameters
  census_interval = 1,
  
  # important experiment parameters
  experiment_params = bind_rows(simulation_type_wild_type(num_simulations = 32),
    simulation_type_knockdown(num_simulations = 0)),
  
  # store the number of reaction firing
  store_reaction_firings = FALSE,
  
  # store the propensities value
  store_reaction_propensities = FALSE,
  
  # GRN activation
  compute_cellwise_grn = FALSE,
  
  # whether or not perform dimensionality reduction
  compute_dimred = TRUE,
  
  # compute propensity ratio
  compute_propensity_ratios = FALSE,
  
  # generate noise to the kinetics of each simulations, takes feature_info and feature_network as input, modified and returns a list
  kinetics_noise_function = kinetics_noise_simple(mean = 1, sd = 0.005))
```

**kinetics_noise_function has two choices**, `kinetics_noise_simple(mean = 1, sd = 0.005)` or `kinetics_noise_none()`:

`mean = 1` the mean level of noise, `sd` standard deviation of noise.
```{r}
?kinetics_noise_none()
```
  
- `experiment_params = experiment_snapshot()` 
library construction and experiment sampling process
```{r}
?experiment_snapshot()
```

In all
```{r}
set.seed(10)
model <- 
  initialise_model(
    num_tfs = 12,
    num_targets = 30,
    num_hks = 15,
    backbone = backbone_bifurcating(),
    verbose = TRUE,
    download_cache_dir = "~/.cache/dyngen",
    num_cores = 8
  )

plot_backbone_statenet(model)
plot_backbone_modulenet(model)
```

## Step 2, generate the model
```
generate_dataset(model, output_dir = NULL, make_plots = FALSE, store_dimred = model$simulation_params$compute_dimred, store_cellwise_grn = model$simulation_params$compute_cellwise_grn, store_propensity_ratios = model$simulation_params$compute_propensity_ratio)
```

Whether or not store propensity ratio, grn, dimred...


```{r}
generate_dataset(
      model,
      # store the result out directly
      output_dir = exp$dataset_folder(id),
      make_plots = TRUE
    )
gc()
```


**Note: more directly, store the model**
if expand the function using `body(generate_dataset)`, then the generation is equal to 
```{r}
model <- model %>% generate_tf_network() %>% generate_feature_network() %>% 
        generate_kinetics() %>% generate_gold_standard() %>% 
        generate_cells() %>% generate_experiment()
```
And it also store the `model` and `dataset` and `plots`

## Step 3, store the result

```{r}
pwalk(
  design_datasets, function(id, seed, backbone_name, tr_rate_multiplier){
    dataset <- read_rds(exp$dataset_file(id))
    spliced <- as.matrix(dataset$expression)
    unspliced <- as.matrix(dataset$expression_unspliced)
    protein <- as.matrix(dataset$expression_protein)
    cell_info <- dataset$cell_info
    write.csv(spliced,file = paste0("./velo_benchmark/",id,"_spliced.csv"))
    write.csv(unspliced,file = paste0("./velo_benchmark/",id,"_unspliced.csv"))
    write.csv(protein,file = paste0("./velo_benchmark/",id,"_unspliced.csv"))
    write.csv(cell_info,file = paste0("./velo_benchmark/",id,"_cell_info.csv"))
  }
  
)
```


## Results
Usually stored as `dataset` and `model`. And the `dataset` is generated from `model`
```{r}
dataset <- wrap_dataset(model, store_dimred = store_dimred, 
        store_cellwise_grn = store_cellwise_grn, store_propensity_ratios = store_propensity_ratios)
```

And `wrap_dataset` **DO LOG TRANSFORM**, main part of the function body
```{r}
dataset <- wrap_expression(id = model$id, counts = model$experiment$counts_mrna, 
        expression = as(log2(model$experiment$counts_mrna + 1), 
            "dgCMatrix"), expression_unspliced = as(log2(model$experiment$counts_premrna + 
            1), "dgCMatrix"), expression_protein = as(log2(model$experiment$counts_protein + 
            1), "dgCMatrix"), cell_info = model$experiment$cell_info %>% 
            select(-from, -to, -time), feature_info = model$experiment$feature_info) %>% 
        add_trajectory(milestone_network = model$gold_standard$network, 
            progressions = model$experiment$cell_info %>% select(cell_id, 
                from, to, percentage = time))
```

**The components of model**
```
 [1] "backbone"               "numbers"                "distance_metric"       
 [4] "tf_network_params"      "feature_network_params" "kinetics_params"       
 [7] "gold_standard_params"   "simulation_params"      "experiment_params"     
[10] "verbose"                "download_cache_dir"     "num_cores"             
[13] "id"                     "feature_info"           "feature_network"       
[16] "simulation_system"      "gold_standard"          "simulations"           
[19] "experiment"    
```
And `model$experiment` mainly stored the raw count and cell info
```
[1] "counts_premrna"    "counts_mrna"       "counts_protein"    "feature_info"     
[5] "cell_info"         "cellwise_grn"      "propensity_ratios"
```

And `model$simulations` mainly stored the propensity results
```
[1] "meta"                  "counts"                "cellwise_grn"         
[4] "reaction_firings"      "reaction_propensities" "propensity_ratios"    
[7] "kd_multiplier"         "perturbed_parameters"  "dimred"               
[10] "dimred_projected" 
```

## Generation step explained
`generate_kinetics()` the body part is explained as below
```{r}
{
    if (model$verbose) 
        cat("Precompiling reactions for simulations\n")
    reactions <- .generate_cells_precompile_reactions(model)
    if (model$verbose) 
        cat("Running ", nrow(model$simulation_params$experiment_params), 
            " simulations\n", sep = "")
    simulations <- pbapply::pblapply(X = seq_len(nrow(model$simulation_params$experiment_params)), 
        cl = model$num_cores, FUN = .generate_cells_simulate_cell, 
        model = model, reactions = reactions)
    
    # true count is obtained from here
    model$simulations <- lst(meta = map_df(simulations, "meta"), 
        counts = do.call(rbind, map(simulations, "counts")), 
        cellwise_grn = do.call(rbind, map(simulations, "cellwise_grn")), 
        reaction_firings = do.call(rbind, map(simulations, "reaction_firings")), 
        reaction_propensities = do.call(rbind, map(simulations, 
            "reaction_propensities")), propensity_ratios = do.call(rbind, 
            map(simulations, "propensity_ratios")), kd_multiplier = do.call(rbind, 
            map(simulations, "kd_multiplier")), perturbed_parameters = do.call(rbind, 
            map(simulations, "perturbed_parameters")))
    if (model$verbose) 
        cat("Mapping simulations to gold standard\n", sep = "")
    if (!is.null(model[["gold_standard"]])) {
        model$simulations$meta <- .generate_cells_predict_state(model)
    }
    else {
        model$simulations$meta <- model$simulations$meta %>% 
            rename(sim_time = time)
    }
    if (model$simulation_params$compute_dimred) {
        if (model$verbose) 
            cat("Performing dimred\n", sep = "")
        model <- model %>% calculate_dimred()
    }
    model
}
```
`generate_cells()` the body part is explained as below, mainly change `model$simulations`
```{r}
{
    if (model$verbose) 
        cat("Precompiling reactions for simulations\n")
    reactions <- .generate_cells_precompile_reactions(model)
    if (model$verbose) 
        cat("Running ", nrow(model$simulation_params$experiment_params), 
            " simulations\n", sep = "")
            
    # main simulation process
    simulations <- pbapply::pblapply(X = seq_len(nrow(model$simulation_params$experiment_params)), 
        cl = model$num_cores, FUN = .generate_cells_simulate_cell, 
        model = model, reactions = reactions)
    
    # true results stored into model simulations, prior to expriment sampling process
    # lst(...), similar to base::list() but for tibble, for details: https://tibble.tidyverse.org/reference/lst.html
    model$simulations <- lst(meta = map_df(simulations, "meta"), 
    
        # map extract count(col) element from simulations, rbind bind them together as count
        counts = do.call(rbind, map(simulations, "counts")), 
        cellwise_grn = do.call(rbind, map(simulations, "cellwise_grn")), 
        reaction_firings = do.call(rbind, map(simulations, "reaction_firings")), 
        
        # reaction propensities is NULL, weird
        reaction_propensities = do.call(rbind, map(simulations, 
            "reaction_propensities")), propensity_ratios = do.call(rbind, 
            map(simulations, "propensity_ratios")), kd_multiplier = do.call(rbind, 
            map(simulations, "kd_multiplier")), perturbed_parameters = do.call(rbind, 
            map(simulations, "perturbed_parameters")))
            
    if (model$verbose) 
        cat("Mapping simulations to gold standard\n", sep = "")
    if (!is.null(model[["gold_standard"]])) {
    
        # state prediction, here must for trajectories inference, we only have one true order.
        model$simulations$meta <- .generate_cells_predict_state(model)
    }
    
    # if gold standard doesn't exist, then change sim_time in meta to time and use it,
    # here meta must store "true time"
    else {
        model$simulations$meta <- model$simulations$meta %>% 
            rename(sim_time = time)
    }
    if (model$simulation_params$compute_dimred) {
        if (model$verbose) 
            cat("Performing dimred\n", sep = "")
        model <- model %>% calculate_dimred()
    }
    model
}
```

`generate_experiment()` the body part is explained as below

**Note that the true count is stored in `tsim_counts<-model$simulations$counts`**
```{r}
{
    if (model$verbose) 
        cat("Simulating experiment\n")
    step_ixs <- .generate_experiment_sample_cells(model)
    
    # drop = FALSE, good habbit when indexing the data.frame
    cell_info <- model$simulations$meta[step_ixs, , drop = FALSE] %>% 
        mutate(step_ix = step_ixs)
    if ("group" %in% names(attributes(step_ixs))) {
        cell_info$cell_group <- attr(step_ixs, "group")
    }
    else {
        cell_info <- cell_info %>% sample_n(n())
    }
    cell_info <- cell_info %>% mutate(cell_id = paste0("cell", 
        row_number())) %>% select(cell_id, step_ix, simulation_i, 
        sim_time, from, to, time, everything())
    step_ixs <- cell_info$step_ix
    
    # this is the true count
    tsim_counts <- model$simulations$counts[step_ixs, , drop = FALSE]
    
    # generate_experiment_fetch_realcount download real counts from website
    # used to guide the sampling process
    realcount <- .generate_experiment_fetch_realcount(model)
    
    # sampling process
    count_simulation <- .simulate_counts_from_realcounts(tsim_counts, 
        realcount, cell_info, sample_capture_rate = model$experiment_params$sample_capture_rate)
        
    # sampling result   
    sim_counts <- count_simulation$sim_counts
    cell_info <- count_simulation$cell_info
    mol_info <- count_simulation$mol_info
    
    # store the result, count w: pre_mrna, x: mrna, y:protein
    sim_wcounts <- sim_counts[, model$feature_info$mol_premrna, 
        drop = FALSE]
    sim_xcounts <- sim_counts[, model$feature_info$mol_mrna, 
        drop = FALSE]
    sim_ycounts <- sim_counts[, model$feature_info$mol_protein, 
        drop = FALSE]
    dimnames(sim_wcounts) <- dimnames(sim_xcounts) <- dimnames(sim_ycounts) <- list(cell_info$cell_id, 
        model$feature_info$feature_id)
    if (model$simulation_params$compute_cellwise_grn) {
        sim_cellwise_grn <- model$simulations$cellwise_grn[step_ixs, 
            , drop = FALSE]
        rownames(sim_cellwise_grn) <- cell_info$cell_id
    }
    else {
        sim_cellwise_grn <- NULL
    }
    
    # store propensity ratio or not
    if (model$simulation_params$compute_propensity_ratios) {
        sim_propensity_ratios <- model$simulations$propensity_ratios[step_ixs, 
            , drop = FALSE]
        rownames(sim_propensity_ratios) <- cell_info$cell_id
    }
    else {
        sim_propensity_ratios <- NULL
    }
    
    # Store the results, should extract real count(sampled in experiment) from model$experiment 
    model$experiment <- list(counts_premrna = sim_wcounts, counts_mrna = sim_xcounts, 
        counts_protein = sim_ycounts, feature_info = model$feature_info, 
        cell_info = cell_info, cellwise_grn = sim_cellwise_grn, 
        propensity_ratios = sim_propensity_ratios)
    model
}
```

## Extract true count before library construction
When dyngen draw reaction process, it use true count rather than sampled version

```{r}
# model is a list 
model <- read_rds("temporary_files/usecase_rna_velocity_b2b/datasets/bifurcating_1_1_1/model.rds")
print(names(model))
print(names(model$simulations))
```



```{r}

# model$feature_info$mol_premrna is a character vector store the feature(genes) name with premrna
# other are similar
# true_count is a matrix with colnames as gene names above
step_ixs <- cell_info$step_ix
true_counts <- model$simulations$counts[step_ixs, , drop = FALSE]
unspliced_t <- true_counts[, model$feature_info$mol_premrna, drop = FALSE]
spliced_t <- true_counts[, model$feature_info$mol_mrna, drop = FALSE]
protein_t <- true_counts[, model$feature_info$mol_protein, drop = FALSE]
```